---
title: "Updated GAMs with BE method"
author: Eric Scott
format: 
  html:
    toc: true
execute: 
  echo: false
---

```{r}
#| label: setup
#| include: false
library(targets)
library(here)
library(fs)
library(dplyr)
library(tinytable)
tar_load(c(
#   summary_plot_50, summary_plot_650, summary_plot_2500,
  summary_plot,
  sd_plot,
  count_plot,
  appraisal_50, appraisal_650, appraisal_2500,
  k_check_50, k_check_650, k_check_2500,
  slopes_plot_slopes_gam_50, slopes_plot_slopes_gam_650, slopes_plot_slopes_gam_2500
 ),
 store = here("_targets"))
```

```{r}
#| label: copy-files
#| include: false
dir_create(here("docs/img"))
summary_plot <- 
    file_copy(
        path = here(summary_plot), new_path = path(here("docs", "img"), path_file(summary_plot)),
        overwrite = TRUE
    )

sd_plot <- 
    file_copy(
        path = here(sd_plot), new_path = path(here("docs", "img"), path_file(sd_plot)),
        overwrite = TRUE
    )
count_plot <- 
    file_copy(
        path = here(count_plot), new_path = path(here("docs", "img"), path_file(count_plot)),
        overwrite = TRUE
    )

appraisals_orig <- c(appraisal_50, appraisal_650, appraisal_2500)
appraisals <- 
    file_copy(
        path = here(appraisals_orig), new_path = path(here("docs", "img"), path_file(appraisals_orig)),
        overwrite = TRUE
    )

slope_plots_orig <- c(slopes_plot_slopes_gam_50, slopes_plot_slopes_gam_650, slopes_plot_slopes_gam_2500)
slope_plots <- 
    file_copy(
        path = here(slope_plots_orig), new_path = path(here("docs", "img"), path_file(slope_plots_orig)),
        overwrite = TRUE
    )
```

For now, I've only run this with three thresholds: 50, 650, and 2500 GDD (ºF), but I've also written the code in a way that it is easy to add additional thresholds.
Currently runs on Jetstream2 instance with 16 cores and 60GB RAM

## Summary statistics plots

### Min, Mean, Max

::: callout-note
These are calculated with `na.rm = TRUE` to remove NAs, so not every pixel necessarily has the same number of years used to calculate these summary statistics.
:::

```{r}
knitr::include_graphics(summary_plot)
```

### Standard Deviation

```{r}
knitr::include_graphics(sd_plot)
```


### Number of years threshold is reached

These plots may be helpful along with standard deviation for interpretation since standard deviation may be calculated with a different "sample size" for different pixels.  (not really sample size, since `NA`s aren't really missing—just indicate the threshold was never reached)

```{r}
knitr::include_graphics(count_plot)
```

## Model Diagnostics

Model diagnostics produced with `gratia::appraise()`.
This indicates that some GDD thresholds might need "bespoke" treatment.
E.g. residuals for 2500 are leptokurtic and might be better if a student t error distribution was used instead of gaussian.
On the other hand, $n$ is so high it probably makes no difference.

```{r}
knitr::include_graphics(appraisals)
```

Similarly, the number of basis functions used (`k'`) might need to be adjusted for some models. 
Ideally `edf` (estimated degrees of freedom) should be much less than `k'`, which is not the case for the spatial smooth (`ti(x,y)`) for 650 GDD.

```{r}
df <- dplyr::bind_rows(k_check_50, k_check_650, k_check_2500)
tt(df %>% select(-gam)) %>% 
    group_tt(i = df$gam) %>%
    style_tt(i = .@group_index_i, background = "#b2b2b2")
```

## Average Slopes Plots

These plots average instantaneous slopes at each year.  At each pixel, the hypothesis that the slope = 0 is tested and resulting p-values are corrected for false discovery rate.  Actual p-values are not displayed, they're just used to "block out" areas that are not significantly different from zero.

::: callout-important
These color bars ***do have*** the same range.
:::



```{r}
knitr::include_graphics(slope_plots)
```

::: callout-warning
For the 2500 GDD threshold, there actually isn't any data in the most northern parts of the US and the fitted values from the GAM should be ignored there.  A refined version of this plot could remove any pixels too far from where there is actual observed data (a common practice for visualizing 2D smooths from GAMs).
:::